### 2022-06-27 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_ncbitaxonomy_softwarenote@9e37654d5cf3da8de7bf1978556b9f3a5e26afeb 🚀

All these considerations are actually important when matching species
names both within and across datasets. Let us consider the following
species survey of individual fishes, European chub, \emph{Cyprinus
cephalus}, \emph{Leuciscus cephalus}, \emph{Squalius cephalus}: all are
the same species (\emph{S. cephalus}), referred to as one of the
vernacular (European chub) and two formerly accepted names now
classified as synonyms. A cautious estimate of diversity based on the
user-supplied names would give \(n=4\) species, when there is in fact
only one.

A package with the ability to handle the sources of errors outlined
above, and especially while provide an authoritative classification, can
accelerate the work of consuming large volumes of biodiversity data. For
example, this package was used in the process of developing the
\emph{CLOVER} database (Gibb et al. 2021) of host-virus associations, by
reconciling the names of viruses and mammals from four different
sources, where all of the issues described above were present.

@@

An up-to-date version of the documentation for \texttt{NCBITaxonomy.jl}
can be found online from the EcoJulia documentation hub
\href{https://docs.ecojulia.org/NCBITaxonomy.jl/stable/}{https://docs.ecojulia.org/},
including examples and in-line documentation of every method. The
package is released under the MIT license. Contributions can be made in
the form of issues (bug reports, questions, features suggestions) and
pull requests. The package can be checked out and installed anonymously
from the central Julia repository:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{using}\NormalTok{ Pkg}

\CommentTok{\# This line should go in the Julia configuration file {-} note that the path}
\CommentTok{\# will be created if it doesn\textquotesingle{}t exist, and will be used to store the}
\CommentTok{\# raw taxonomic table}
\ConstantTok{ENV}\NormalTok{[}\StringTok{"NCBITAXONOMY\_PATH"}\NormalTok{] }\OperatorTok{=}\NormalTok{ joinpath(homedir()}\OperatorTok{,} \StringTok{"data"}\OperatorTok{,} \StringTok{"NCBITaxonomy.jl"}\NormalTok{)}

\NormalTok{Pkg.add(}\StringTok{"NCBITaxonomy"}\NormalTok{) }\CommentTok{\# Dowloading the files may take a long time}
\end{Highlighting}
\end{Shaded}

The package will download the most recent version of the NCBI taxonomy
database, and transform in into a set of Apache Arrow files ready for
use. Note that the \texttt{NCBITAXONOMY\_PATH} can specified on a
per-project basis, and as long as the package is not re-built, the local
set of tables downloaded from NCBI will not change; this way, users can
re-run an analysis with a guarantee that the underlying taxonomic
backbone has not changed.

@@

Name finding is primarily done through the \texttt{taxon} function,
which admits either a unique NCBI identifier (\emph{e.g.}
\texttt{taxon(36219)} for the bogue \emph{Boops boops}), a string
(\texttt{taxon("Boops\ boops")}), or a data frame with a restricted list
of names (see the next section). The \texttt{taxon} method has
additional arguments to perform fuzzy matching in order to catch
possible typos (\texttt{taxon("Boops\ bops";\ strict=false)}), to
perform a lowercase search (useful when alphanumeric codes are part of
the taxon name, like for some viruses), and to restrict the the search
to a specific taxonomic rank.

@@

Depending on the operating system, either of these three names can be
returned; compare to the output of a case insensitive name search:

@@

The \texttt{taxon} function will either return a \texttt{NCBITaxon}
object (made of a \texttt{name} and \texttt{id}), or throw either a
\texttt{NameHasNoDirectMatch} (with instructions about how to possible
solve it, using the \texttt{similarnames} function), or a
\texttt{NameHasMultipleMatches} (listing the possible valid matches, and
suggesting to use \texttt{alternativetaxa} to find the correct one).
Therefore, the common way to work with the \texttt{taxon} function would
be to wrap it in a \texttt{try}/\texttt{catch} statement:

@@

These functions will not demand any user input in the form of key
presses (though they can be wrapped in additional code to allow it), as
they are intended to run on clusters without supervision. The
\texttt{taxon} function has good scaling using muliple threads. For
convenience in rapidly getting a taxon for demonstration purposes, we
also provide a string macro, whereby \emph{e.g.}
\texttt{ncbi"Procyon\ lotor"} will return the taxon object for the
raccoon.

@@

As the full NCBI names table has over 3 million entries at the time of
writing, we have provided a number of functions to restrict the scope of
names that are searched. These are driven by the NCBI \emph{divisions}.
For example \texttt{nf\ =\ mammalfilter(true)} will return a data frame
containing the names of mammals, inclusive of rodents and primates, and
can be used with \emph{e.g.} \texttt{taxon(nf,\ "Pan")}. This has the
dual advantage of making search faster, but also of avoiding matching on
names that are shared by another taxonomic group (which is not an issue
with \emph{Pan}, but is an issue with \emph{e.g.} \emph{Io} as mentioned
in the introduction).

Note that the use of a restricted list of names can have significant
performance consequences: compare, for example, the time taken to return
the taxon \emph{Pan} (ID 9596) in the entire database, in all mammals,
and in all primates:

@@

Clearly, the optimal search strategy is to (i) rely on name filters to
ensure that search are conducted within the appropriate NCBI division,
and (ii) only rely on fuzzy matching when the strict or lowercase match
fails to return a name, as fuzzy matching can result in order of
magnitude more run time and memory footprint. These numbers were
obtained on a single Intel i7-8665U CPU (@ (1.90GHz). Using
\texttt{"chimpanzees"} as the search string (the NCBI recognized
vernacular for \emph{Pan}) gave qualitatively similar results,
suggesting that there is no performance cost associated with working
with synonyms or verncular input data.

@@

In order to facilitate working with names, we provide the
\texttt{authority} function (gives the full taxonomic authority for a
name), \texttt{synonyms} (to get alternative valid names),
\texttt{vernacular} (for English common names), and \texttt{rank} (for
the taxonomic rank).

@@

The \texttt{children} function will return all nodes that are directly
descended from a taxon; the \texttt{descendants} function will
recursively apply this function to all descendants of these nodes, until
only terminal leaves are reached. The \texttt{parent} function is an
``upwards'' equivalent, giving that taxon from which a taxon descents;
the \texttt{lineage} function chains calls to \texttt{parent} until
either \texttt{taxon(1)} (the taxonomy root) or an arbitrary ancestor is
reached.

The \texttt{taxonomicdistance} function (and its in-place equivalent,
\texttt{taxonomicdistance!}, which uses memory-efficient re-allocation
if the user needs to change the distance between taxonomic ranks) uses
the Shimatani (2001) approach to reconstruct a matrix of distances based
on taxonomy, which can serve as a rough proxy when no phylogenies are
available.

@@

>>>>

All these considerations become important when matching species names
both within and across datasets. Let us consider the hypothetical
species survey of riverine fishes: European chub, \emph{Cyprinus
cephalus}, \emph{Leuciscus cephalus}, \emph{Squalius cephalus}. All are
the same species (\emph{S. cephalus}), referred to as one of the
vernacular (European chub) and two formerly accepted names now
classified as synonyms (but still present in the litterature). A
cautious estimate of diversity based on the user-supplied names would
give \(n=4\) species, when there is in fact only one. When the size of
biodiversity datasets increases, and notably when the taxonomic scope of
these datasets explodes, including organisms for which ``names'' are a
fuzzier concept (for example, \emph{Influenza A virus
(A/Sydney/05/97-like(H3N2))} is a valid name for a common influenza
strain, although one that lacks a taxonomic rank), the feasibility of
manual curation decreases.

In this manuscript, we describe \texttt{NCBITaxonomy.jl}, a Julia
package that provides advanced name matching and error handling
capacities for the reconciliation of taxonomic names to the NCBI
database. This package was used to facilitate the development of the
\emph{CLOVER} (Gibb et al. 2021) database of host-virus associations, by
reconciling the names of viruses and mammals from four different
sources, where all of the issues described above were present. More
recently, it has become part of the automated curation of data for the
\emph{VIRION} (Carlson et al. 2022) database, which automatically
curates an up-to-date, authoritative host-mammal network from dozens of
heterogeneous sources. We describe the core capacities of this package,
and highlight how it enables safe, high-performance name reconciliation.

@@

An up-to-date version of the documentation for \texttt{NCBITaxonomy.jl}
can be found online from the EcoJulia documentation hub
\href{https://docs.ecojulia.org/NCBITaxonomy.jl/stable/}{https://docs.ecojulia.org/},
including examples and in-line documentation of every method. The
package is released under the MIT license. Contributions can be made in
the form of issues (bug reports, questions, features suggestions) and
pull requests.

In order to achieve good performance, the package will first retrieve
the latest (as validated by its checksum) NCBI taxonomy backbone, store
it locally, and pre-process it as a set of Julia data tables. By
default, the taxonomy will be downloaded to the Julia package
repository, which is not an ideal solution, and therefore we recommend
that users set an environment variable to specificy where the data will
be loaded from (this path will be created if it doesn't exist):

\begin{Shaded}
\begin{Highlighting}[]
\ConstantTok{ENV}\NormalTok{[}\StringTok{"NCBITAXONOMY\_PATH"}\NormalTok{] }\OperatorTok{=}\NormalTok{ joinpath(homedir()}\OperatorTok{,} \StringTok{"data"}\OperatorTok{,} \StringTok{"NCBITaxonomy.jl"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Note that this location can be different for different projects, as the
package is able to update the taxoonmic backbone (and will indeed prompt
the user to do so if the taxonomy is more than 90 days old). The package
can then be checked out and installed anonymously from the central Julia
repository:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{using}\NormalTok{ Pkg}
\NormalTok{Pkg.add(}\StringTok{"NCBITaxonomy"}\NormalTok{) }\CommentTok{\# Dowloading the files may take a long time}
\end{Highlighting}
\end{Shaded}

As long as the package is not re-built, the local set of tables
downloaded from NCBI will not change; this way, users can re-run an
analysis with a guarantee that the underlying taxonomic backbone has not
changed, which is not the case when relying on API queries. In order to
update the taxonomic backbone, users can call the \texttt{build}
function of Julia's package manager (\texttt{{]}build\ NCBITaxonomy}),
which will download the most recent version of all files.

@@

Name finding is primarily done through the \texttt{taxon} function,
which admits either a unique NCBI identifier (\emph{e.g.}
\texttt{taxon(36219)} for the bogue \emph{Boops boops}), a string
(\texttt{taxon("Boops\ boops")}), or a data frame with a restricted list
of names in order to create a name finder function (see the next
section). The \texttt{taxon} method has additional arguments to perform
fuzzy matching in order to catch possible typos
(\texttt{taxon("Boops\ bops";\ strict=false)}), to perform a lowercase
search (useful when alphanumeric codes are part of the taxon name, like
for some viruses), and to restrict the the search to a specific
taxonomic rank. The \texttt{taxon} function also accepts a
\texttt{preferscientificname} keyword, to prevent matching vernacular
names; the use of this keyword ought to be informed by knowledge about
how the data were entered.

@@

Depending on the operating system (and specifically whether it is
case-sensitive), either of these three names can be returned; compare to
the output of a case insensitive name search:

@@

When it succeeds, \texttt{taxon} will return a \texttt{NCBITaxon} object
(made of a \texttt{name} string field, and an \texttt{id} numerical
field). That being said, the package is designed under the assumption
that ambiguities should yield an error for the user to handle. There are
two such errors: \texttt{NameHasNoDirectMatch} (with instructions about
how to possible solve it, using the \texttt{similarnames} function), or
a \texttt{NameHasMultipleMatches} (listing the possible valid matches,
and suggesting to use \texttt{alternativetaxa} to find the correct one).
Therefore, the common way to work with the \texttt{taxon} function would
be to wrap it in a \texttt{try}/\texttt{catch} statement:

@@

These functions will not demand any user input in the form of key
presses (though they can be wrapped in additional code to allow it), as
they are intended to run on clusters or virtual machines without
supervision. The \texttt{taxon} function has good scaling using muliple
threads. For convenience in rapidly getting a taxon for demonstration
purposes, we also provide a string macro, whereby \emph{e.g.}
\texttt{ncbi"Procyon\ lotor"} will return the taxon object for the
raccoon.

@@

As the full NCBI names table has over 3 million entries at the time of
writing, we have provided a number of functions to restrict the scope of
names that are searched. These are driven by the NCBI \emph{divisions}.
For example \texttt{nf\ =\ mammalfilter(true)} will return a data frame
containing the names of mammals, inclusive of rodents and primates, and
can be used with \emph{e.g.} \texttt{taxon(nf,\ "Pan")}. This has the
dual advantage of making search faster, but also of avoiding matching on
names that are shared by another taxonomic group (which is not an issue
with \emph{Pan}, but is an issue with \emph{e.g.} \emph{Io} as mentioned
in the introduction, or with the common name \emph{Lizard}, which
fuzzy-matches on the hemipteran genus \emph{Lisarda} rather than the
class \emph{Lepidosauria}).

Note that the use of a restricted list of names can have significant
performance consequences: compare, for example, the time taken to return
the taxon \emph{Pan} in the entire database, in all mammals, and in all
primates:

@@

Clearly, the optimal search strategy is to (i) rely on name filters to
ensure that search are conducted within the appropriate NCBI division,
and (ii) only rely on fuzzy matching when the strict or lowercase match
fails to return a name, as fuzzy matching can result in order of
magnitude more run time and memory footprint. These numbers were
obtained on a single Intel i7-8665U CPU (@ (1.90GHz). Using
\texttt{"chimpanzees"} as the search string (one of the NCBI recognized
vernaculars for \emph{Pan}) gave qualitatively similar results,
suggesting that there is no performance cost associated with working
with synonyms or verncular input data.

@@

In order to facilitate working with names, we provide the
\texttt{authority} function (gives the full taxonomic authority for a
name), \texttt{synonyms} (to get alternative valid names),
\texttt{vernacular} (for English common names), and \texttt{rank} (for
the taxonomic rank). These functions are not used in name matching, but
are often useful in the post-processing of results.

@@

The \texttt{children} function will return all nodes that are directly
descended from a taxon; the \texttt{descendants} function will
recursively apply this function to all descendants of these nodes, until
only terminal leaves are reached. The \texttt{parent} function is an
``upwards'' equivalent, giving the taxon from which a taxon descends;
the \texttt{lineage} function chains calls to \texttt{parent} until
either \texttt{taxon(1)} (the taxonomy root) or an arbitrary ancestor is
reached.

The \texttt{taxonomicdistance} function (and its in-place equivalent,
\texttt{taxonomicdistance!}, which uses memory-efficient re-allocation
if the user needs to change the distance between taxonomic ranks) uses
the Shimatani (2001) approach to reconstruct a matrix of distances based
on taxonomy, which can serve as a rough proxy when no phylogenies are
available. This allows coarse estimations of taxonomic diversity based
on species lists. The default distance between taxoonmic levels is as in
Shimatani (2001) (\emph{i.e.} species have a distance of 0, genus of 1,
family of 2, sbu-classes of 3, and everything else 4), but specific
scores can be passed for \emph{any} taxonomic level know to the NCBI
name table.

@@

\leavevmode\hypertarget{ref-Carlson2022GloVir}{}%
Carlson, C. J. et al. 2022. The Global Virome in One Network (VIRION):
An Atlas of Vertebrate-Virus Associations. - mBio in press.
